---
title: "SSB 2020 Workshop: Using the Open Tree of Life for your Research"
subtitle: "Get a dated tree for your _taxa of interest_ (using R) with `datelife` and the Open Tree of Life"
author: "Luna L. Sánchez-Reyes"
date: "`r Sys.Date()`"
output:
    pdf_document
header-includes: \usepackage{graphicx}
graphics: yes
vignette: >
  %\VignetteIndexEntry{datelife package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include = FALSE, warnings = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ">",
  eval = FALSE
)
library("datelife")
data("opentree_chronograms")
runtime <- attr(opentree_chronograms, "running_time")
# for provisional testing:
start_time <- Sys.time()
Sys.sleep(10)
end_time <- Sys.time()
runtime <- end_time - start_time
```

# Requirements for this tutorial

From R, install the packages `rotl`, `ape`, `devtools` and `stringr` with the function `install.packages()`, 
and the package `datelife` with the function `install_github()`. 
Load them into your workspace with `library()` or `require()`. 
If you do not want to load the packages, you can call functions specifying their package using two colons and the syntax `package_name::function_name()`. 
This implies more typing, but gives more clarity to reproduce the workflow later. So we will use that syntax for this tutorial.
An exception to this are functions from packages that are "preloaded" (such as `library()` form `base` or `install.packages()`, from `utils`) that can be simply called by their name.

```{r}
install.packages(c("rotl", "ape", "devtools", "stringr"))
library(rotl)
library(devtools)
library(stringr)
devtools::install_github("phylotastic/datelife")
library(datelife)
```

# Find your taxa in the Open Tree of Life Taxonomy

Get the _ott ids_ for your taxa with `tnrs_match_names()`. This will generate a "match_names" object.
```{r, eval = TRUE}
my_taxa <- c("amphibians", "canis", "felis", "delphinidae", "spheniscidae")
resolved_names <- rotl::tnrs_match_names(my_taxa)
class(resolved_names)
resolved_names
```

There are two ways to extract the _ott ids_ from a "match_names" object:
```{r}
rotl::ott_id(resolved_names) # as a list
resolved_names$ott_id # as a vector
```
# Get an induced subtree for your set of taxa

```{r, eval = TRUE, results = 'hide'}
mytree <- rotl::tol_induced_subtree(resolved_names$ott_id)
```

What does this warning mean?
  
  
```{r, eval = TRUE, results= 'asis', warning= FALSE, error = TRUE}
ape::plot.phylo(mytree, cex = 1)
```

This is cool! 
  
Now, what if you want a synthetic tree containing some or all descendants from your taxa of interest?

# Get a subtree of one taxon

We can extract a subtree of all descendants of one taxon at a time using the function `tol_subtree()` and the amphibians _ott id_. 

Try to extract a subtree of all amphibians.
Get the _ott id_ first. It is already stored in the `resolved_names` object, but you can run the function `tnrs_match_names()` again if you want.
```{r, eval = TRUE}
amphibia_ott_id <- resolved_names$ott_id[1] # extract the ott id from resolved_names
# amphibia_ott_id <- rotl::tnrs_match_names("amphibians")$ott_id # OR run tnrs again
```

Now, extract the subtree from the Open Tree synthetic tree using `tol_subtree()`.
```{r, eval = TRUE, results= 'hide', warning= FALSE}
amphibia_subtree <- rotl::tol_subtree(ott_id = amphibia_ott_id)
```
```{r, eval = TRUE}
amphibia_subtree
```
This is a large tree!

Try to extract a subtree for the genus _Canis_. It should be way smaller!

```{r, eval = TRUE}
canis <- rotl::tnrs_match_names("canis")  # get the ott id of the genus Canis (dogs)
canis
```
```{r, eval = TRUE, results= 'hide', warning= FALSE, error = TRUE}
subtree <- rotl::tol_subtree(canis$ott_id) # get the subtree
```

Why am I getting an error saying that the _node id_ was not found and that my taxon is "broken"??
  
Generally, this happens when phylogenetic information does not match taxonomic information. 
For example, extinct lineages are sometimes phylogenetically included within the taxon but are taxonomically excluded, making the taxon appear as paraphyletic.
On the browser, you can still get to the subtree. From R, you need to do something else first.

# Get a subtree of a "broken" taxon

There is a way to find out that the group is "broken" before trying to get the subtree and getting an error.
```{r, eval = TRUE, results= 'hide', warning= FALSE, error = TRUE}
rotl::is_in_tree(canis$ott_id)
```
We say that a taxon is not in the tree when its _ott id_ is not assigned to a node in the synthetic tree.
This is the reason why we get an error when we try to get a subtree using the _ott id_.

Then, how do I get a subtree of my "broken" taxon?
There are some options.

## a) Get a subtree using the _node id_ instead of the _ott id_
`rotl` has a function that gets for you all info from the node containing a taxon. That includes the actual _node id_.
```{r, eval = TRUE, warning= FALSE, error = TRUE}
canis_node_info <- rotl::tol_node_info(canis$ott_id)
canis_node_info
```
The _node_ that contains _Canis_ is "`r canis_node_info$node_id`". We can use it to get a subtree with `tol_subtree()`
```{r, eval = TRUE, results= 'hide', warning= FALSE, error = TRUE}
canis_node_subtree <- rotl::tol_subtree(node_id = canis_node_info$node_id)
```
```{r, eval = TRUE, results= 'asis', warning= FALSE, error = TRUE, fig.height = 9}
ape::plot.phylo(canis_node_subtree, cex = 0.5)
```

Nice! We got a subtree of 85 tips, containing all descendants from the node that also contains _Canis_.

This includes species assigned to genera other than _Canis_. 

What if I _really, really_ want to get a tree containing species within the genus _Canis_ only?
  
## b) Get an induced subtree of taxonomic children

We can get the _ott ids_ of the taxonomic children of our taxon of interest and use the function `tol_induced_subtree()`.

First, get the taxonomic children.
```{r, eval = TRUE, results= 'hide', warning= FALSE, error = TRUE}
canis_taxonomy <- rotl::taxonomy_subtree(canis$ott_id)
```
```{r, eval = TRUE, results= 'asis', warning= FALSE, error = TRUE}
canis_taxonomy
```
Now, extract the _ott ids_.
```{r, eval = TRUE, results= 'hide', warning= FALSE, error = TRUE}
canis_taxonomy_ott_ids <- datelife::extract_ott_ids(x = canis_taxonomy$tip_label)
```
Try to get an induced subtree of _Canis_ taxonomic children.

```{r, eval = TRUE, results = 'hide', warning= FALSE, error = TRUE}
canis_taxonomy_subtree <- rotl::tol_induced_subtree(canis_taxonomy_ott_ids)
```
It is often not possible to get an induced subtree of all taxonomic children from a taxon, 
because some of them are not in the tree.

To verify which ones are giving us trouble, we can use the function `is_in_tree()` again.
```{r, eval = TRUE, warning= FALSE, error = TRUE, results = 'hide'}
canis_in_tree <- sapply(canis_taxonomy_ott_ids, rotl::is_in_tree)
canis_taxonomy_ott_ids_intree <- canis_taxonomy_ott_ids[canis_in_tree]
canis_taxonomy_subtree <- rotl::tol_induced_subtree(canis_taxonomy_ott_ids_intree)
```
```{r, eval = TRUE, results= 'asis', warning= FALSE, error = TRUE, fig.height = 6}
ape::plot.phylo(canis_taxonomy_subtree, cex = 0.5)
```
There! We have a synthetic subtree (derived from phylogenetic information) containing only the taxonomic children of _Canis_.

It is sometimes thought that flags can help to detect taxa that are not in the tree, but they can be misleading.
Get the flags of _Canis_ taxonomic children.
```{r, eval = TRUE, warning= FALSE, error = TRUE}
canis_taxonomy_ott_ids_taxon_info <- rotl::taxonomy_taxon_info(canis_taxonomy_ott_ids)
canis_taxonomy_ott_ids_taxon_info_flags <- rotl::flags(canis_taxonomy_ott_ids_taxon_info)
in_tree <- sapply(canis_taxonomy_ott_ids, rotl::is_in_tree)
flags_not_in_tree <- canis_taxonomy_ott_ids_taxon_info_flags[!in_tree]
flags_not_in_tree <- unique(unlist(flags_not_in_tree))
flags_in_tree <- canis_taxonomy_ott_ids_taxon_info_flags[in_tree]
flags_in_tree <- unique(unlist(flags_in_tree))
```
```{r, eval = TRUE, warning= FALSE, error = TRUE}
flags_not_in_tree 
flags_in_tree
```

Some terms such as "infraspecific" and "extinct" are used to flag both taxa that are **not** in the tree and taxa that **are** in the tree.

# Get a subtree of a rank

We can use the function `tax_rank()` to extract all children at or below the species level.
First, get all ranks.
```{r, eval = TRUE, results= 'hide', warning= FALSE, error = TRUE}
canis_ranks <- rotl::tax_rank(canis_taxonomy_ott_ids_taxon_info)
```
Then, get your desired rank only.
```{r, eval = TRUE}
canis_species_index <- canis_ranks %in% "species"
canis_species_in_tree <- canis_taxonomy_ott_ids[canis_species_index & canis_in_tree]
```

Get an induced subtree of these _ott ids_.
```{r, eval = TRUE, results= 'hide', warning= FALSE, error = TRUE}
canis_species_subtree <- rotl::tol_induced_subtree(canis_species_in_tree)
```
```{r, eval = TRUE}
canis_species_subtree
```
```{r, eval = TRUE, results= 'asis', warning= FALSE, error = TRUE, fig.height = 6}
ape::plot.phylo(canis_species_subtree, cex = 0.8)
```

What if we want to get all taxa of a taxonomic rank above the species level?
Try to get all amphibian families.

Extract ott ids from tip labels of subtree of all amphibians.
```{r, eval = TRUE, results= 'hide', warning= FALSE, error = TRUE}
amphibia_subtree_ott_ids <- datelife::extract_ott_ids(x = amphibia_subtree$tip.label)
```
Get their taxon information. This might take about 10 minutes or so.
```{r, eval = FALSE}
amphibia_ott_ids_taxon_info <- rotl::taxonomy_taxon_info(amphibia_subtree_ott_ids)
```

We already know that all children of amphibia are in the tree. So we only need to extract the ones belonging to our desired rank.
```{r, eval = FALSE}
amphibia_ranks <- rotl::tax_rank(amphibia_subtree_ott_ids_taxon_info)
unique(amphibia_ranks)
# [[1]]
# [1] "species"
# 
# [[2]]
# [1] "subspecies"
# 
# [[3]]
# [1] "no rank - terminal"
```

We have implemented a function in `datelife` to extract all taxa of a given rank. It is also sligthly faster.

```{r, eval = TRUE, results= 'hide', warning= FALSE, error = TRUE}
amphibia_families <- datelife::get_ott_children(ott_ids = amphibia_ott_id, ott_rank = "family")
```
Now, get the subtree of families only.
```{r, eval = TRUE, results= 'hide', warning= FALSE, error = TRUE}
amphibia_families_subtree <- rotl::tol_induced_subtree(amphibia_families$Amphibia$ott_id)
```
```{r, eval = TRUE, results= 'asis', warning= FALSE, error = TRUE, fig.height = 8}
amphibia_families_subtree
ape::plot.phylo(amphibia_families_subtree, cex = 0.5)
```
Awesome!
  
# Get branch length information (proportional to time) for you taxa

You can explore the studies supporting the synthetic tree, and find out which ones have branch lengths.

In the package `datelife`, we have implemented a workflow that extracts all studies containing information from at least two taxa.
It only works at the species level.

```{r, eval = TRUE}
apes <- c("Pongo", "Pan", "Gorilla", "Hoolock", "Homo")
resolved_names <- rotl::tnrs_match_names(apes)
```

```{r, eval = TRUE, results = 'hide'}
apes_dr <- datelife::get_datelife_result(resolved_names$unique_name, get_spp_from_taxon = TRUE)
```
```{r, eval = TRUE}
names(apes_dr)[1]
apes_dr[[1]] # look at the first element of the list
names(apes_dr)[length(apes_dr)]
apes_dr[[length(apes_dr)]] # look at the last element of the list
```

We have now a list of matrices storing time of lineage divergence data for all taxon pairs.
Lists are named by the study citation, so we have that information handy at all times.

We can summarize it.
```{r, eval = TRUE, results = 'hide', message = FALSE}
apes_phylo_median <-  datelife::summarize_datelife_result(apes_dr,summary_format = "phylo_median")
```
```{r, eval = TRUE, results= 'asis', warning= FALSE, error = TRUE, fig.height = 6}
apes_phylo_median
ape::plot.phylo(apes_phylo_median, cex = 1)
ape::axisPhylo()
````